---
title: "Google Sheets API setup using Google Cloud"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Google Sheets API setup using Google Cloud}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A goal of this package is to enable the use of Google Sheets for collaborative data entry with 
familiar spreadsheet features.    The package could also have used MS Office on-line excel editing, but excel files 
on desktop have historically had issues with datetime conversion (Mac vs Windows), line endings, and character encoding. 

A main feature then is to read in google 'sheets' files as data for validation and processing.   However, accessing google drive files requires significant setup for an 'app' to access files.   This vignette describes how that works. 

This package heavily depends on the great work from Posit in the `googlesheets4` package, which in turn relies on the Posit-authored [gargle](Bryan J, Citro C, Wickham H 2023). gargle: Utilities for Working with Google APIs. https://gargle.r-lib.org.) package

However before even starting R you need a Google Cloud project.   The gargle package has some instructions for this 

`vignette("get-api-credentials", package = "gargle")`

This is essentially a desktop app to read private google sheets, so in that guide we wan at to create "OAuth 2.0 client ID and secret"

The 'gargle' vignette says 
>Note that most users of gargle-using packages do not need to read this and can just enjoy the automatic token flow. This article is for people who have a specific reason to be more proactive about auth.

But I've found that one get several "insufficient permissions" at my institution to access files on a shared drive without going through this process. 





*DRAFT* summary:

- get a google workspace account
- get a google cloud account.   If your team is all using the same google workspace (e.g. at the same university or company) it's much easier and  you should create a google cloud account in that domain.  Most institutions restrict this so you may need help from your institutions cloud gatekeepers.   Often it's the IT department.   You can create a free account which requires a credit card, but this process will not have any charges and you will have to use 

- create a new cloud project in the console
- in the console go to the APIs page https://console.cloud.google.com/apis/
- create 'credentials' which is aka "service account" 
  - https://console.cloud.google.com/apis/credentials
  - click 'create credentials' and follow the prompts. 
    - give the credential a name (which would be the name of this package)
- setup the OAuth consent screen.  If you are all close collaborators you may wonder why you have to setup a consent screen but it's google and you do
  - give it name related to your project that is unique but that all collaborators would instantly recognize, perhaps "ProjectX-R" 
  - it's much easier to make this consent 'internal' which limits users in the same 'workspace' as you (my institution has essential one workspace for all users ).  If using external collaborators with email/google drive accounts that are not your institution pick 'external.'  this will change what you need to do but these notes don't cover that. 
  
  - google also calls these 'consent screens' *"apps"* so you have to give it an app name - use the same as the credentials name above.  
  - user support email is your email
  - app logo -  you can skip all of that
  - App domain: if you are working in a lab or interdisciplinary project, use that domain, or the domain of some department in your institution that you are a member of or is helpful for this kind of thing, like "biology.myuni.edu" or "x-project.biology.myuni.edu"
  - "Authorized domains" - if you are using internal app, then put the lowest domain which is typically the domain on your institutional email address (e.g. "myuni.edu" or "myuni.ac.uk" or "my-non-profit.org")
     - since this is just for internal collaborators, use the same url for all of home page, privacy policy link and terms of service link.   
  - If using external users, you may have to add a domain for each user that is their instution, or perhaps use gmail for everyone.  (not tested)
  
  
- enable APIs
  - to use the google sheets package, enable the sheets api.  it may be necessary to enable the google drive API (which I did)

  - set for read-only
    - there is a 'scope' setting and may want to change to ti 'read only' since this package only reads google drive files.  when setting up the apis, use this scope if possible.  
    
- download credentials

In the OAuth 2.0 Client IDs, download the JSON file needs to be accessible by your R session.  It should not be inside your R code folder which is a git repository and this identity file should never be checked into git (for security).  It does not go into the package itself, and must be given to each person using the package.   To make it flexible, update the `.Renviron` file to point to that file. It may need to be the full path to the file.   For example
`PROJECT_AUTH_FILE='/Users/myuserid/downloads/client_secret_77etc_blahblah.apps.googleusercontent.com.json'




commands to try: 

```R
drive_auth_json_file <- 'path/to/json/file.json'
# may want to check that file exists 
googlesheets4::googlesheets4::gs4_auth_configure(path=drive_auth_json_file)
googlesheets4::gs4_auth(email='myname@institution.whatever', scopes="drive.readonly")
```
  

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(commruleR)
```
