---
title: "Validating Community Assembly Rules Project Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating Community Assembly Rules Project Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
require(dplyr)
library(commruleR)
```

## Load data specific to the Comm Assembly Rules project stored in this package

```{r, start, echo=TRUE}
# Validate package rules
data(commassembly_rules_biomass_str)
data(commassembly_rules_env_str)

# list of URLS from google drive to a data sheet
doc_with_list_url <- Sys.getenv('TEST_ID_LIST_URL')
id_column = 'ID_new'
urls.df <- read_url_list(gurl = doc_with_list_url, id_column = id_column, url_column='url')
print(paste(nrow(urls.df), "urls to read"))
```

Read in the first URL as a test.  There will be warnings if the columns are not the correct type or name
```{r set_example_url, echo=TRUE}
# get the example from one of the URLS

# Randome sheet
test_gsheet_url <- sample(urls.df$url, 1)
# OR set the example sheet manually
# test_gsheet_url <- 'https://docs.google.com/spreadsheets/d/1Npcre4y4LnzIU_4v_vqJiVyzLtXGZ1lpYGRnXcfi0X8/edit?gid=0#gid=0'

print(paste("We'll use this test sheet:", test_gsheet_url))


```


```{r, read_one_biomass, echo=TRUE}
# read in one, and validate the column names
# this will throw warnings if there are invalid column names
biomass.df <- read_data_sheet(gurl = test_gsheet_url, 
                              tab_name = 'biomass_data', 
                              spec.df = commassembly_rules_biomass_str)

biomass.df
```

env sheet

```{r, read_one_env, echo=TRUE}
env.df <- read_data_sheet(gurl = test_gsheet_url, 
                          tab_name = 'env_data', 
                          spec.df = commassembly_rules_env_str)
head(env.df)
```

## Validation

the above does validate the column headers and types.  (TDB problems output from readr)

Let's try using the validation rules

```{r, validate biomass from file, echo=TRUE, error=TRUE}

biomass_validation_file <- '../inst/rules/biomass_validation_rules.yaml'
file.exists(biomass_validation_file)
validation_summary <- validate::summary(validate_from_file(biomass.df, biomass_validation_file))
validation_summary

```

Validation: Just show the fails
```{r, show validation fails}
if(sum(validation_summary$fails) == 0) {
  print("Validation Passed, return true")
} else {
  fails <- validation_summary[validation_summary$fails > 0,]
  fails
}

```

**Validate the 'env' tab**

```{r, validate env from file, echo=TRUE, error=TRUE}

env_validation_file <- '../inst/rules/env_validation_rules.yaml'
# file.exists(env_validation_file)
validation_summary <- validate::summary(validate_from_file(env.df, env_validation_file))
if(sum(validation_summary$fails) == 0) {
  print("Validation Passed, return true")
} else {
  print("Validation fails")
  fails <- validation_summary[validation_summary$fails > 0,]
  fails
}

```


## Function to read, check columns, validate and save the CSVS of these two tabs

```{r, functions to write sheets, echo=TRUE, error=TRUE}
csv_folder = '../L0'
dir.create(csv_folder, showWarnings = FALSE)
biomass_validation_file <- '../inst/rules/biomass_validation_rules.yaml'
env_validation_file <- '../inst/rules/env_validation_rules.yaml'

# sheet_data is one row of the URLS data frame, as a list, url is ..$url
save_csvs<- function(sheet_data){
  print(paste(sheet_data$id, sheet_data$url))
  url <- sheet_data$url
  tryCatch({
    env.df <- read_data_sheet(url, tab_name = 'env_data', spec.df = commassembly_rules_env_str)
    }, error=function(e){ 
      print(paste("reading or columns error"))
      print(e)
      return (c(NA, NA))
    }
  )
    
  if(!exists('env.df')) {
    print(paste("could not read sheet: "))
    return (c(NA, NA))
  }
  
  tryCatch({
    biomass.df <- (read_data_sheet(url, tab_name = 'biomass_data', spec.df = commassembly_rules_biomass_str))
    }, error=function(e){ 
      print(paste("reading or columns error"))
      print(e)
      return (c(NA, NA))

    }
  )
  
  if(!exists('biomass.df')) {
    print(paste("could not read sheet "))
    return (c(NA, NA))
  } 
  
  biomass_validation_summary <- validate::summary(validate_from_file(biomass.df, biomass_validation_file))
  if(sum(validation_summary$fails) == 0) {
      print("Validation Passed, return true")
  } else {
    fails <- dplr::select(validation_summary, fails > 0)
    print(fails)
}
  
  id_new = sheet_data$ID_new
  biomass_file_name <- file.path(csv_folder, paste0('biomass_', id_new, '.csv'))
  write.csv(biomass.df, biomass_file_name, row.names = FALSE)
  env_file_name <- file.path(csv_folder, paste0('env_', id_new, '.csv'))
  write.csv(env.df, env_file_name, row.names = FALSE)
  
  return(c(biomass_file_name, env_file_name))
}


```

###  Save one to a csv and output the filename

```{r, save_one, echo=TRUE, error=TRUE}

test_url_num <- sample(1:nrow(urls.df), 1)
test_gsheet <- as.list(urls.df[test_url_num,])
test_file_names<- save_csvs(test_gsheet)

print(test_file_names)
```

### Try to read the CSV back in


```{r}
test_df <- read_data_csv(test_file_names[1], spec.df = commassembly_rules_biomass_str)
test_df
```


```{r, save them all, echo=TRUE, error=TRUE}

# before running this, authenticate

if(!gsheet_auth_setup(drive_email = Sys.getenv('PROJECT_EMAIL'))) {
  warning('could not authenticate to google sheets')
} else {
  for(i in seq(1:nrow(urls.df))) {
    print(urls.df$id[i])
    tryCatch({
      study_info <- as.list(urls.df[i,])
      file_names<- save_csvs(study_info)
      print(file_names)
      }, error=function(e) print(e))
  }
}
```

### Build a database

```{r, combine, echo=TRUE}
# getting the list of CSV file paths is flexible so that some could be 
# excluded or multiple folders combined

biomass_files <- dir('../L0', pattern = "biomass.*\\.csv", 
                          full.names = TRUE, include.dirs = TRUE)
biomass.df <- aggregate_csvs(csv_list=biomass_files, spec.df=commassembly_rules_biomass_str)

# some of these are not saving or reading correctly
#env_files <- dir('../L0', pattern = "env.*\\.csv", full.names = TRUE, include.dirs = TRUE)
# env.df <- aggregate_csvs(csv_list=env_files, spec.df=commassembly_rules_env_str)

print(paste('biomass rows', nrow(biomass.df))) # , " env rows", nrow(env.df)))
```
