---
title: "Validating Community Assembly Rules Project Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating Community Assembly Rules Project Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
require(dplyr)
library(commruleR)
```

## Authenticate to gdrive/sheets.  Requires an email account

```{r gdrive setup, echo=TRUE}

## EDIT AND ADD YOUR EMAIL IF YOU DON'T HAVE AN EMAIL IN .Renviron FILE
# if you have not setup the .Renvfile, or would like to test a different email 
# address, set it here.   If none is sent, the auth setup function will look in 
# the environment (e.g. .Renviron file)
drive_email <- NULL
commruleR::gsheet_auth_setup(drive_email = drive_email)
```



## Load data specific to the Comm Assembly Rules project stored in this package

```{r, start, echo=TRUE}
# Validate package rules
data(commassembly_rules_biomass_str)
data(commassembly_rules_env_str)

# list of URLS from google drive to a data sheet
doc_with_list_url <- Sys.getenv('TEST_ID_LIST_URL')
id_column = 'ID_new'
urls.df <- read_url_list(gurl = doc_with_list_url, id_column = id_column, url_column='url')
print(paste(nrow(urls.df), "urls to read"))
```

Read in the first URL as a test.  There will be warnings if the columns are not the correct type or name
```{r set_example_url, echo=TRUE}
# get the example from one of the URLS

# Randome sheet
test_gsheet_url <- sample(urls.df$url, 1)
# OR set the example sheet manually
# test_gsheet_url <- 'https://docs.google.com/spreadsheets/d/1Npcre4y4LnzIU_4v_vqJiVyzLtXGZ1lpYGRnXcfi0X8/edit?gid=0#gid=0'

print(paste("We'll use this test sheet:", test_gsheet_url))


```


```{r, read_one_biomass, echo=TRUE}
# test of basic reading, and validate the column names
# this will throw warnings if there are invalid column names

biomass.df <- read_data_sheet(gurl = test_gsheet_url, 
                              tab_name = 'biomass_data', 
                              spec.df = commassembly_rules_biomass_str)

biomass.df
```

env sheet

```{r, read_one_env, echo=TRUE}
env.df <- read_data_sheet(gurl = test_gsheet_url, 
                          tab_name = 'env_data', 
                          spec.df = commassembly_rules_env_str)
head(env.df)
```

## Validation

the above does validate the column headers and types.  (TDB problems output from readr)

Let's try using the validation rules

```{r, validate biomass from file, echo=TRUE, error=TRUE}

biomass_validation_file <- '../inst/rules/biomass_validation_rules.yaml'
file.exists(biomass_validation_file)
validation_summary <- validate::summary(validate_from_file(biomass.df, biomass_validation_file))
validation_summary

```

Validation: Just show the fails
```{r, show validation fails}
if(sum(validation_summary$fails) == 0) {
  print("Validation Passed, return true")
} else {
  fails <- validation_summary[validation_summary$fails > 0,]
  fails
}

```

**Validate the 'env' tab**

```{r, validate env from file, echo=TRUE, error=TRUE}

read_data_sheet_save_warnings <- errorSaver(read_data_sheet)
env.df <- read_data_sheet_save_warnings(gurl = test_gsheet_url, 
                              tab_name = 'env_data', 
                              spec.df = commassembly_rules_env_str)

if("warnings" %in% names(env.df)) { 
  print(env.df$warnings)
  # get just the data
  env.df <- env.df[[1]]
}

```

```{r}
env_validation_file <- '../inst/rules/env_validation_rules.yaml'
# file.exists(env_validation_file)
validation_summary <- validate::summary(validate_from_file(env.df, env_validation_file))
if(sum(validation_summary$fails) == 0) {
  print("Validation Passed, return true")
} else {
  print("Validation fails")
  fails <- validation_summary[validation_summary$fails > 0,]
  fails
}

```

### find missing column error

this is example code to only show validation errors and not to save anything

disabled for now, see loop to  validate below

```{r echo=FALSE}
# before running this, authenticate. 
# if you don't have ths PROJECT_EMAIL set in .Renviron, edit the next line
# drive_email = Sys.getenv('PROJECT_EMAIL')
# stopifnot(gsheet_auth_setup(drive_email = drive_email))
# 
# env_validation_file <- '../inst/rules/env_validation_rules.yaml'
# googledrive::local_drive_quiet()
# 
# for(i in seq(1:nrow(urls.df))) {
#     data_sheet_url <- as.list(urls.df[i,])
#     print(paste(data_sheet_url$who, data_sheet_url$id, data_sheet_url$url))
#     
#     # this will create list with 'warnings' if there are warnings
#     
#     read_data_sheet_save_warnings <- commruleR::errorSaver(read_data_sheet)
#     env.df <- read_data_sheet_save_warnings(gurl = test_gsheet_url, 
#                               tab_name = 'env_data', 
#                               spec.df = commassembly_rules_env_str)
# 
#     if("warnings" %in% names(env.df)) { 
#       print(env.df$warnings)
#       # get just the data
#       env.df <- env.df[[1]]
#     }
# 
#     # env.df <- read_data_sheet(data_sheet_url$url, 
#     #                             tab_name = 'env_data', 
#     #                             spec.df = commassembly_rules_env_str)
# 
#     
#     if('warnings' %in% names(env.df)){
#       print(data_sheet_url)
#       print(env.df)
#       
#     # } else {
#       env_validation_summary <- validate::summary(validate_from_file(env.df, env_validation_file))
#       if(sum(validation_summary$fails) != 0) {
#         fails <- dplry::filter(validation_summary, fails > 0)
#         print(paste(fails$name, "has", fails$fails, "invalid rows"))
#       }
#     }
# }

```


## Function to read, check columns, validate and save the CSVS of these two tabs

```{r, functions to write sheets, echo=FALSE, error=TRUE}
csv_folder = '../L0'
dir.create(csv_folder, showWarnings = FALSE)
biomass_validation_file <- '../inst/rules/biomass_validation_rules.yaml'
env_validation_file <- '../inst/rules/env_validation_rules.yaml'

# sheet_data is one row of the URLS data frame, as a list, url is ..$url
read_and_report<- function(url, tab_name, spec.df){
  read_data_sheet_save_warnings <- commruleR::errorSaver(read_data_sheet)
  data.df <- read_data_sheet_save_warnings(gurl = url, 
                              tab_name, 
                              spec.df)

  if("warnings" %in% names(data.df)) { 
      print(paste("reading warnings for tab_name:", tab_name, "url", url))
      print(data.df$warnings)
      # get just the data so what's left can be validated
      data.df <- data.df[[1]]
  }
    
  return(data.df)
}

validation_report<- function(data.df, validation_file){
    confrontation<- validate_from_file(data.df, validation_file)
    validation_summary <- validate::summary(confrontation)
    if(sum(validation_summary$fails) == 0) {
      return(TRUE)
    } else {
      print(validation_summary)
      #fails <- dplyr::filter(validation_summary , fails > 0)
      #print(fails)
      return(FALSE)
  }
}

save_csvs<- function(sheet_data){
  # print(paste(sheet_data$id, sheet_data$url))

  url <- sheet_data$url
  
  #### ENV  
  env.df <- read_and_report(url, 
                            tab_name = 'env_data', 
                            spec.df = commassembly_rules_env_str)
  if(!exists('env.df')) {
    print(paste("could not read sheet: "))
    return (c(NA, NA))
  } 
  env_valid <- validation_report(env.df, env_validation_file)

  ## BIOMASS READ
  biomass.df <- read_and_report(url, 
                                tab_name = 'biomass_data', 
                                spec.df = commassembly_rules_biomass_str)
  if(!exists('biomass.df')) {
    print(paste("could not read sheet "))
    return (c(NA, NA))
  } 
  # BIOMASS VALIDATION
  biomass_valid <- validation_report(biomass.df, biomass_validation_file)
   
  if(biomass_valid && env_valid) {
    # SAVE
    id_new <- sheet_data$ID_new
    biomass_file_name <- file.path(csv_folder, paste0('biomass_', id_new, '.csv'))
    write.csv(biomass.df, biomass_file_name, row.names = FALSE)
    env_file_name <- file.path(csv_folder, paste0('env_', id_new, '.csv'))
    write.csv(env.df, env_file_name, row.names = FALSE)
    return(c(biomass_file_name, env_file_name))
  } else {
    print(paste("did not validate biomass", biomass_valid, " env ", env_valid ))
    return(c(NA,NA))
  }
}


```

###  Save one to a csv and output the filename

```{r, save_one, echo=TRUE, error=TRUE}

test_url_num <- sample(1:nrow(urls.df), 1)
test_gsheet <- as.list(urls.df[test_url_num,])
print(test_gsheet)
test_file_names<- save_csvs(test_gsheet)

print(test_file_names)
```

### Try to read the CSV back in


```{r}
test_df <- read_data_csv(test_file_names[1], spec.df = commassembly_rules_biomass_str)
test_df
```

## Loop that reads all URLS in the url list. 

This is complete code, with duplicates from above, to show how to validate all CSVs

```{r, save them all, echo=TRUE, error=TRUE}

data(commassembly_rules_biomass_str)
data(commassembly_rules_env_str)

# before running this, authenticate. 
# if you don't have ths PROJECT_EMAIL set in .Renviron, edit the next line
drive_email = Sys.getenv('PROJECT_EMAIL')
stopifnot(gsheet_auth_setup(drive_email = drive_email))


# list of URLS from google drive to a data sheet
doc_with_list_url <- Sys.getenv('TEST_ID_LIST_URL')
id_column = 'ID_new'
urls.df <- read_url_list(gurl = doc_with_list_url, id_column = id_column, url_column='url')

for(i in seq(1:nrow(urls.df))) {
  print(paste(urls.df$ID_new[i], urls.df$id[i], urls.df$who[i]))
  tryCatch({
    study_info <- as.list(urls.df[i,])
    file_names<- save_csvs(study_info)
    print(file_names)
    }, error=function(e) print(e))
}
```



### Build a database

```{r, combine, echo=TRUE}
# getting the list of CSV file paths is flexible so that some could be 
# excluded or multiple folders combined

biomass_files <- dir('../L0', pattern = "biomass.*\\.csv", 
                          full.names = TRUE, include.dirs = TRUE)
biomass.df <- aggregate_csvs(csv_list=biomass_files, spec.df=commassembly_rules_biomass_str)

# some of these are not saving or reading correctly
#env_files <- dir('../L0', pattern = "env.*\\.csv", full.names = TRUE, include.dirs = TRUE)
# env.df <- aggregate_csvs(csv_list=env_files, spec.df=commassembly_rules_env_str)

print(paste('biomass rows', nrow(biomass.df))) # , " env rows", nrow(env.df)))
```
